---
title: "Lab 5: Time Series and a Map"
author: "Ilayda Dinc"
date: "2/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
# having message false and warning false in the global options applies it to all.

```{r}
# For general stuff:
library(tidyverse)
library(janitor)
library(lubridate)
library(here)
library(paletteer)

# For ts stuff: 
library(tsibble)
library(fable)
library(fabletools)
library(feasts)
library(forecast)

# For spatial stuff: 
library(sf)
library(tmap)
library(mapview)



```

## Monthly US energy consumption (renewables)
```{r}
us_renew <- read_csv(here("data", "renewables_cons_prod.csv")) %>% 
  clean_names()
```


in the console unique(us_renew$description)


Make the description all lower case
only keep observation where "description" variable contains "consumption"
remove any observations where " description variable contains "total
```{r}
renew_clean <- us_renew %>% 
  mutate(description = str_to_lower(description)) %>%  #we wanna override the existing column
  filter(str_detect(description, pattern = "consumption")) %>% 
  filter(!str_detect(description, pattern = "total"))


```

### Convert `yyyymm` column to date with `lubridate`

```{r}
renew_date <- renew_clean %>% 
  mutate(yr_mo_day = lubridate::parse_date_time(yyyymm, "ym")) %>% 
  mutate(month_sep = yearmonth(yr_mo_day)) %>% #coerce to tsibble `yearmonth` format
  mutate(value = as.numeric(value)) %>% 
  drop_na(month_sep, value)

# Want to parse the year and month? We may use this later...
renew_parsed <-renew_date %>% 
  mutate(month = month(yr_mo_day, label = TRUE)) %>% 
  mutate(year = year(yr_mo_day))
```

```{r}
renew_gg <- ggplot(data = renew_date, aes(x = month_sep, y = value, group = description)) +
  geom_line(aes(color = description)) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 350))

renew_gg
```

Now try updating your color palette using options from paletteer. Use `View(palettes_d_names)` to see all of the discrete scale options. We'll want a palette with at least 7 colors (length >= 7). Find a name of a palette that you like, then update your graph by adding `scale_color_paletteer_d("package::palette")`. Like, if I want to use the `calecopal::figmtn` palette, 

d indicates discrete
I'd add:

`renew_gg + scale_color_paletteer_d("calecopal::figmtn")`

Try some out!

Updating colors with paletteer palettes: 

```{r}
renew_gg +
  scale_color_paletteer_d("nationalparkcolors::GeneralGrant")
```
```{r}
renew_gg +
  scale_color_paletteer_d("calecopal::figmtn")
```

### Coerce to a tsibble:

```{r}
renew_ts <- as_tsibble(renew_parsed, key = description, index = month_sep) # you don't have to specify the key here but sometimes it is needed
```


### Look at the data in a few different ways:
```{r}
renew_ts %>% autoplot(value) # tell it what variable you wanna look at over time
renew_ts %>% gg_subseries(value) # to break this up
#renew_ts %>% gg_season(value)   # did not work

# What if gg_season() didn't work? Well we can make this with ggplot anyway!
# Remember our other parsed version (renew parsed):

ggplot(data = renew_parsed, aes(x = month, y = value, group = year)) +
  geom_line(aes(color = year)) +
  facet_wrap(~ description, 
             ncol = 1, 
             scales = "free", 
             strip.position = "right")  # i want only one column, I want strip to be on the right. you can update the color scheme later on.
  
```


### Get just the hydroelectric energy consumption data:
```{r}
hydro_ts <- renew_ts %>% 
  filter(description == "hydroelectric power consumption")

# Explore: 
hydro_ts %>% autoplot(value)
hydro_ts %>% gg_subseries(value)

#hydro_ts %>% gg_season(value)

# OK, what if gg_season() doesn't work?
# It's just a function that uses ggplot() to do things we already know how to do in ggplot()!
ggplot(hydro_ts, aes(x = month, y = value, group = year)) +
  geom_line(aes(color = year))
  
```


### Calculate summary data by time using `index_by()`

What if we want to calculate consumption by quarter? 

We'll use `index_by()` to tell R which "windows" to calculate a value with in. 



Quarterly:
```{r}
hydro_quarterly <- hydro_ts %>% 
  index_by(year_qu = ~ yearquarter(.)) %>% # monthly aggregates
  summarise(
    avg_consumption = mean(value)
  )

head(hydro_quarterly)
```

Or annually: 
```{r}
hydro_annual <- hydro_ts %>% 
  index_by(annual = ~year(.)) %>% 
  summarize(
    avg_consumption = mean(value)
  )

ggplot(data = hydro_annual, aes(x = annual, y = avg_consumption)) +
  geom_line(color = "darkorange") +
  geom_smooth(color = "purple",
              size = 0.2,
              linetype = "dashed",
              fill = "purple",
              alpha = 0.2) +
  theme_minimal()
```

And if you have higher interval data (e.g. hourly), then you can calculate summaries by week, month, etc. using functions from `tsibble` like: 

- `yearweek`
- `yearmonth`

### Decompose the hydro consumption ts data

First, let's check the decomposition (STL):
```{r}
# Find STL decomposition
dcmp <- hydro_ts %>%
  model(STL(value ~ season(window = 5)))

# View the components
# components(dcmp)

# Visualize the decomposed components
components(dcmp) %>% autoplot() +
  theme_minimal()

# Let's check out the residuals:
hist(components(dcmp)$remainder)

```


### Explore the ACF

```{r}

hydro_ts %>% 
  ACF(value) %>% 
  autoplot()

```
observation that are 12 months apart are more correlated
We see peaks at 12 months: annual-difference similarities in consumption. 


## DANGER DANGER

### Forecast future hydro power consumption 

```{r}
hydro_model <- hydro_ts %>%
  model(
    arima = ARIMA(value),
    ets = ETS(value)
  ) %>%
  fabletools::forecast(h = "4 years")

hydro_model %>% 
  autoplot(filter(hydro_ts, 
                  year(month_sep) > 2010), 
           level = NULL)

```

## Map-of-the-day

A world map with bubbles!
```{r}

# Get spatial data: 
world <- read_sf(dsn = here("data","TM_WORLD_BORDERS_SIMPL-0.3-1"), layer = "TM_WORLD_BORDERS_SIMPL-0.3") %>% clean_names()

# Quick & easy option to see those polygons (also for points, lines!)
mapview(world)

# ggplot (static)
world_base <- ggplot(data = world) +
  geom_sf(aes(fill = pop2005),
          color = NA) + 
  scale_fill_paletteer_c("viridis::viridis") +
  theme_minimal()

world_base

# Let's crop it: 
world_base +
  coord_sf(xlim = c(-20, 50), ylim = c(-40, 40), expand = FALSE)
```

## Making this a web page:

- Knit to create the html
- Push all updates back to GitHub

Since you are working in a gh-pages branch, this can be a webpage! 

- Go to your-github-username.github.io/repo-name/file-name-prefix
- You should see the webpage containing your knitted document!

- Troubleshooting:
        - Don't including a trailing slash (404 error)
        - Don't include the .Rmd extension (will ask to download .Rmd)
        
# End Lab Week 5
